Effectiveness of BNT162b2 booster doses in England: a observational study in OpenSAFELY

### 2022-02-22 manuscript/draft-manuscript.Rmd
edits from feedback

---
title: "Effectiveness of COVID-19 Booster doses in England: a sequential trials study in OpenSAFELY"
output:
  word_document: default
  html_document:
    keep_md: yes
    self_contained: yes
  bookdown::html_document2:
    number_sections: no
    toc: no
  pdf_document:
    toc: no
bibliography: references.bib
csl: vancouver.csl
link-citations: yes
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

@@

@@

hazards_none <- read_csv(fs::path(output_dir_os, "none", "effects.csv")) %>%
  mutate(
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(number_format(accuracy=0.01)(hr), " (", number_format(accuracy=0.01)(hr.ll), "-", number_format(accuracy=0.01)(hr.ul), ")")
  )

hazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv")) %>%
  mutate(
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(number_format(accuracy=0.01)(hr), " (", number_format(accuracy=0.01)(hr.ll), "-", number_format(accuracy=0.01)(hr.ul), ")")
  )

#cmlinc3 <- cmlinc %>% filter(model==3)
#riskdiff3 <- riskdiff %>% filter(model==3)
hazards3_none <- hazards_none %>% filter(model==3)
hazards3_vax12_type <- hazards_vax12_type %>% filter(model==3)

@@

William J Hulme^1^, Elizabeth J Williamson^2^, Amelia Green^1^, Elsie Horne^3,4^, Helen Curtis^1^, Linda Nab^1^, Milan Wiedemann^1^, Ruth Keogh^2^, Edward Parker^2^, Venexia Walker^3,4^, Jonathan AC Sterne^3,4^, Miguel Hernán^6,7^, Ben Goldacre^1^ + OpenSAFELY collaborative.

@@

With the approval of NHS England we used the OpenSAFELY-TPP database, covering 40% of GP practices in England and linked to national coronavirus surveillance, hospital episodes, and death registry data, to estimate the effectiveness of a booster dose in "fully vaccinated" people using a sequential trials approach. On each day of the recruitment period (`r format(as.Date(study_dates$studystart_date), "%d %B %Y")` to `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")`) people receiving a booster dose on that day were recruited to the treatment arm, with a matched sample of unboosted people recruited to the control arm. Controls were matched on geogrpahical region, JCVI priority group, and the week and type of the second vaccine dose. Trial arms across each day were pooled and analysed using a Cox model, controlling for additional confounders. Primary outcomes were recorded SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, unplanned COVID-19-related hospital admission, and COVID-19-related death occurring within 12 weeks of recruitment.

@@

We matched a total of `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` BNT162b2 booster recipients to unboosted controls with only two doses. During the 28-41 days the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(hazards_none, outcome=="postest", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for a positive test, `r filter(hazards_none, outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 hospital admission, and `r filter(hazards_none, outcome=="coviddeath", treatment=="pfizer", model==3, term_left==28) %>% pull(hr_display)` for COVID-19 death.

@@

@@

The national COVID-19 vaccination programme in the United Kingdom began a new phase in September 2021 with the administration of the first booster COVID-19 vaccine doses in those who had already received their primary vaccination course (that is, "fully vaccinated" with a complete first and second dose). Initially offered to those at highest risk of severe disease, eligibility was extended to the entire adult population by the end of the year.

We set out to estimate the effectiveness of booster doses amongst the fully vaccinated adult population in England using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population.

In the vast majority of cases, BNT162b2 and mRNA-1273 vaccines were used, and administered without consideration of the type of primary course. BNT162b2 was used from 16 September 2021, and mRNA-1273 from 29 October 2021 onwards. Consequently, available follow-up of BNT162b2 recipients is longer, and so we restrict the analysis to the effectiveness of BNT162b2boosting only. An analysis of mRNA-1273 booster effectiveness is planned when sufficient follow-up has accrued.

@@

The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at GP practices that use TPP SystmOne electronic health record software. It includes pseudonymized data such as coded diagnoses, medications and physiological parameters. No free text data are included. This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS). HCW status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

## Study design and population

We used a sequential trials approach designed to emulate a sequence of daily two-arm parallel randomised controlled trials. For each trial, the treatment arm comprised people who received a BNT162b2 booster dose on that day, and the control arm comprised the same number of randomly selected unboosted people matched on key characteristics, as described below. This closely follows the approach taken in previous vaccine effectiveness studies in Israeli registry data @dagan2021 @barda2021

Trial recruitment was considered between `r format(as.Date(study_dates$studystart_date), "%d %B %Y")` and `r format(as.Date(study_dates$studyend_date), "%d %B %Y")` inclusive. Trial eligibility on each day of recruitment was as follows: adults (18 and over) who were registered at a GP practice using TPP's SystmOne clinical information system; received a primary COVID-19 vaccination course of either two BNT162b2 doses or two ChAdOx1-S doses (mRNA-1273 was not considered due to small numbers); not a vaccinated health care worker, not a resident in a care or nursing home, and not medically housebound or receiving end-of-life care (due to distinct clustering of vaccine uptake and infection risk in these groups that could not be reasonably measured and controlled for); no evidence of SARS-CoV-2 infection or COVID-19 disease within 90 days prior to recruitment; not undergoing an unplanned hospital admission; information on sex, ethnicity, deprivation, and geographical region was complete.

Additional eligiblity criteria based on sufficiently many people being vaccinated on each day within region, priority group, and second dose week strata were also used to ensure unboosted people were both eligible and able to receive a booster dose at the time of potential recruitment. The trial sequence was terminated early if nobody was eligible for the treatment arm or fewer than 10% of eligible treated people were successfully matched. See supplementary materials for further details.

The control arms were selected using exact 1:1 matching without replacement. Participants were matched on primary course vaccine type (BNT162b2 or ChAdOx1-S), the calendar week of the second vaccine dose, geographical region, prior infection, immunosuppression, and vaccine priority group (as determined by the Joint Committee for Vaccine and immunisation expert working group). People selected as controls were not eligible to be included again as a control in a subsequent trial, but were eligible for selection into the treatment arm of a subsequent trial. Any unmatched booster recipients were dropped.

After exclusions for missing values on demographic variables, there were no missing values in remaining variables as they were each defined by the presence or absence of clinical codes or events in the patient record.

## Exposures, outcomes, covariates

<!--# COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc. -->

The following three outcomes were considered: positive SARS-CoV-2 test; unplanned COVID-19 hospital admission; COVID-19 death. Positive SARS-CoV-2 tests were identified using SGSS records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection. Unplanned COVID-19 hospital admissions were identified using HES in-patient hospital records with U07.1 or U07.2 reason for admission ICD-10 codes. COVID-19 deaths with the same COVID-19 ICD-10 codes mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.

<!--# COVID-19 related A&E attendance and critical/intensive care admission we also considered but there were doubts over data the reliability of event ascertainment in these cases. -->

Participant characteristics used describe trial participants and for confounder adjustment include: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintiles), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" classification, as per national prioritisation guidelines, the number of SARS-CoV-2 tests in the 6 months prior to the study start date (via SGSS), rurality (urban conurbation, urban city or town, rural town or village), evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation), learning disabilities, severe mental illness, undergoing an unplanned hospital admission, undergoing a planned hospital admission. All characteristics were ascertained as at the trial recruitment date.

@@

Time zero was the trial recruitment day, and trial participants were followed up for no more than 10 weeks from this day. Follow-up was censored earlier than this at `r format(as.Date(study_dates$studyend_date), "%d %B %Y")`, death, or GP practice de-registration. Additionally for each matched pair, follow-up for both the boosted and unboosted person was censored if the control received a booster dose.

Trials were pooled and analysed using Cox regression with period-specific hazard ratios for treatment, in two-week periods. Baseline hazards were allowed to differ for participants differing by recruitment day, region, and week and type of second vaccine dose, via Cox stratification. Other variables were included as covariates. A robust variance estimator was used to account for participant clustering.

[The primary estimand is the average treatment effect in the recruited population. To estimate this, expected cumulative incidence rates are computed for each participant assuming treatment was received by everyone in both trial arms at time zero. These curves are then averaged over all participants to give the marginal cumulative incidence. This is repeated assuming nobody is treated in either arm, and their difference taken. Calculating standard errors for these quantities is challenging, whether approached analytically (due to complex variance components) or computationally (via bootstrapping, due to the large datasets involved) and is there not reported. ]

We additionally estimated booster effectiveness in subgroups defined by primary vaccine course (both BNT162b2 or both ChAdOx1-S) using complete model stratification.

@@

Data management and analyses were conducted in Python 3.8 and R version 4.0.2. All code is available at <https://github.com/opensafely/booster-effectiveness>, and is shared openly for review and re-use under MIT open license. Codelists are available at <https://www.opencodelists.org/>. No person-level data is shared. Any reported figures based on counts below 6 are redacted or rounded for disclosure control.

@@

A total of `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_treated))` BNT162b2 booster doses were administered between `r format(as.Date(study_dates$studystart_date), "%d %B %Y")` and `r format(as.Date(study_dates$studyend_date), "%d %B %Y")` in adults registered at a TPP practice, with `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_eligible))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_eligible_treated))`) eligible for inclusion into the treatment arm. Of these, `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_matched_eligible))`) were successfully matched with unboosted controls (Figure 1).

@@

Participant characteristics at trial recruitment between trial arms (Table 1) were typically well-balanced (necessarily so on matched characteristics), though with some exceptions, for instance unboosted controls had higher levels of deprivation, higher rates of learning disabilities and severe mental illness, and lower SARS-CoV-2 testing frequency prior to the study. These unbalanced characteristics were adjusted for in the Cox regression models.

There were `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(fup_years))` person-years of follow-up across both trial arms (Table 2).

@@

#### Figure 1: Booster vaccination uptake, treatment arm eligiblity and matching success

@@

xmin <- min(matchcoverage$vax3_date )
xmax <- max(matchcoverage$vax3_date )+1
matchcoverage  %>%
  filter(treatment=="pfizer") %>%
  mutate(
    # vax12_type = fct_case_when(
    #   vax12_type == "pfizer-pfizer" ~ "BNT162b2",
    #   vax12_type == "az-az" ~ "ChAdOx1-S",
    #   vax12_type == "moderna-moderna" ~ "mRNA-1273",
    #   TRUE ~ NA_character_
    # ),
    course = fct_case_when(
      vax12_type=="pfizer-pfizer" & treatment=="pfizer" ~ "2x BNT162b2  / 1x BNT162b2",
      vax12_type=="az-az" & treatment=="pfizer" ~ "2 x ChAdOx1-S / 1 x BNT162b2",
      vax12_type=="pfizer-pfizer" & treatment=="moderna" ~ "2x BNT162b2 / 1x mRNA-1273",
      vax12_type=="az-az" & treatment=="moderna" ~ "2x ChAdOx1-S / 1x mRNA-1273",
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot() +
  geom_col(
    aes(
      x=vax3_date+0.5,
      y=cumuln,
      group=fct_rev(status),
      fill=fct_rev(status),
      colour=NULL
    ),
    position=position_stack(),
    alpha=0.8,
    width=1
  )+
  geom_rect(xmin=xmin, xmax= xmax+1, ymin=0, ymax=6, fill="grey", colour="transparent")+
  facet_wrap(vars(course))+
  scale_x_date(
    breaks = unique(lubridate::ceiling_date(matchcoverage$vax3_date, "1 month")),
    limits = c(lubridate::floor_date((xmin), "1 month"), NA),
    labels = scales::label_date("%d/%m"),
    expand = expansion(add=1),
  )+
  scale_y_continuous(
    labels = scales::label_number(accuracy = 1, big.mark=","),
    expand = expansion(c(0, NA))
  )+
  scale_fill_brewer(type="qual", palette="Set2", direction = -1)+
  scale_colour_brewer(type="qual", palette="Set2", direction = -1)+
  labs(
    x="Date",
    y="Cumulative booster vaccines",
    colour=NULL,
    fill=NULL,
    alpha=NULL
  ) +
  theme_minimal()+
  theme(
    axis.line.x.bottom = element_line(),
    axis.text.x.top=element_text(hjust=0),
    strip.text.y.right = element_text(angle = 0),
    axis.ticks.x=element_line(),
    legend.position = "bottom"
  )+
  NULL
cat("  \n\n")
```

@@

Participant characteristics as on the day of recruitment into the treatment or control arm.

@@

#### Table 2: Recruitment, follow-up, and event incidence

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath"),
    treatment == "pfizer"
  ) %>%
  left_join(
    matchsummary %>% select(treatment, n_eligible, prop_matched_eligible),
    by=c("treatment")
  ) %>%
  mutate(
    treatment_decr = paste0(
      treatment_descr, " - ",
      label_number(1, big.mark=",")(n_eligible), "eligible, ",
      label_number(1, big.mark=",")(n_1), "matched (",
      label_percent(0.1)(prop_matched_eligible), ")."
    )
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rr, 
    rrCI
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

    rate_0 = "Incidence",
    rate_1 = "Incidence",

    rr = "Incidence rate ratio",
    rrCI = "95% CI"
  ) %>%
  tab_spanner(
    label = "Boosted",
    columns = ends_with("1")
  ) %>%
  tab_spanner(
    label = "Unboosted",
    columns = ends_with("0")
  ) %>%
  fmt_number(
    columns = c("rate_0", "rate_1", "rr"),
    decimals = 3
  ) %>%
  fmt_missing(
    everything(),
    missing_text="--"
  ) %>%
  cols_align(
    align = "right",
    columns = everything()
  )

@@

Trial participants contributed `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="postest", fup_period=="All") %>% pull(events))` positive SARS-CoV-2 tests, `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="covidadmitted", fup_period=="All") %>% pull(events))` COVID-19 hospital admissions, and `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="coviddeath", fup_period=="All") %>% pull(events))` COVID-19 deaths (Table 2).

>>>>

---
title: "Effectiveness of BNT162b2 booster doses in England: a observational study in OpenSAFELY"
output:
  html_document:
    keep_md: yes
    self_contained: yes
  word_document: default
  bookdown::html_document2:
    number_sections: no
    toc: no
  pdf_document:
    toc: no
bibliography: references.bib
csl: vancouver.csl
link-citations: yes
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

@@

studydurationweeks <- as.numeric(as.duration(study_dates$studyend_date - study_dates$studystart_date +1) , "weeks")
```

@@

hazards_none <- read_csv(fs::path(output_dir_os, "none", "effects.csv"))
hazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv"))

hazards <- 
  bind_rows(hazards_none, hazards_vax12_type) %>%
  mutate(

    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
    hr_display = paste0(label_number(accuracy=0.01)(hr), " (", label_number(accuracy=0.01)(hr.ll), "-", label_number(accuracy=0.01)(hr.ul), ")"),
    ve_display = paste0(label_percent(accuracy=0.1)(ve), " (", label_percent(accuracy=0.1, suffix="")(ve.ll), "-", label_percent(accuracy=0.1, suffix="")(ve.ul), ")"),
  )

#cmlinc3 <- cmlinc %>% filter(model==3)
#riskdiff3 <- riskdiff %>% filter(model==3)
hazards3 <- hazards %>% filter(model==3)

@@

William J Hulme^1^, Elizabeth J Williamson^2^, Amelia Green^1^, Elsie Horne^3,4^, Helen Curtis^1^, Linda Nab^1^, Milan Wiedemann^1^, Ruth Keogh^2^, Edward Parker^2^, Venexia Walker^3,4^, Jonathan AC Sterne^3,4^, Miguel A Hernán^6,7^, Ben Goldacre^1^ + OpenSAFELY collaborative.

@@

With the approval of NHS England we used the OpenSAFELY-TPP database, covering 40% of GP practices in England and linked to national coronavirus surveillance, hospital episodes, and death registry data, to estimate the effectiveness of a booster dose in "fully vaccinated" people. On each day of the recruitment period (`r format(as.Date(study_dates$studystart_date), "%d %B %Y")` to `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")`) each eligible person receiving a BNT162b2 booster dose on that day was recruited to the treatment group, and matched with an unboosted person recruited to the control group. Matching factors were region, JCVI priority group, and the day and type of the second vaccine dose. The treated and control groups across each day were pooled. Primary outcomes were documented SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, unplanned COVID-19-related hospital admission, and COVID-19-related death occurring within 12 weeks of recruitment. Hazard ratios were estimated via a Cox model with additional covariates that were not used as matching factors.

@@

We matched `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` BNT162b2 booster recipients to unboosted controls with only two doses. During days 14-20 after the booster, the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only was `r filter(hazards_none, outcome=="postest", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for documented SARS-CoV-2 infection, `r filter(hazards_none, outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards_none, outcome=="coviddeath", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 death.

@@

<!-- TODO -->

<!-- * full description of booster eligiblity over time -->

<!-- * figure 1 chart with SMD / %-point differences -->

<!-- * flowchart incl matching -->

<!-- * check for moderna references and remove if necessary -->

<!-- * check final recruitment day and include in results -->

<!-- * check if rolling average restrictions are ever used (need to output dataset) -->

<!-- * JCVI group spell out -->

<!-- * references -->

@@

The national COVID-19 vaccination programme in England began a new phase on 16 September 2021 with the administration of the first booster COVID-19 vaccine doses in those who had already received their primary vaccination course (a completed first and second dose) (ref <https://www.england.nhs.uk/2021/09/nhs-begins-covid-19-booster-vaccination-campaign/>). Initially offered to those at highest risk of severe disease, eligibility was progressively extended and by 15 December 2021 every adult was eligible (<https://www.england.nhs.uk/2021/12/nhs-booster-bookings-open-to-every-eligible-adult/>). Booster doses were initially available no earlier than six months after dose two, but this was reduced to three months on 8 December 2021, following concerns over the surge in cases and the emergence of the Omicron variant (ref: <https://www.england.nhs.uk/2021/12/people-40-and-over-to-get-their-lifesaving-booster-jab-three-months-on-from-second-dose/>)

<!-- 13 december 30-39 become eligible (https://www.england.nhs.uk/2021/12/nhs-to-roll-out-life-saving-booster-jab-to-people-aged-30-plus/) -->

<!-- 15 december everyone eligible https://www.england.nhs.uk/2021/12/nhs-booster-bookings-open-to-every-eligible-adult/ -->

[context]

We set out to estimate the effectiveness of BNT162b2 mRNA Pfizer-BioNTech booster doses versus two doses only in adults in England using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population. An analysis of effectiveness of Moderna mRNA-1273 boosters, which were administered from 29 October 2021, is planned when sufficient follow-up has accrued.

@@

The OpenSAFELY-TPP database (<https://opensafely.org>) covers 24 million people registered at GP practices that use TPP SystmOne electronic health record software. It includes pseudonymized data such as coded diagnoses, medications and physiological parameters. No free text data are included. This primary care data is linked (via NHS numbers) with A&E attendance and in-patient hospital spell records via NHS Digital's Hospital Episode Statistics (HES), national coronavirus testing records via the Second Generation Surveillance System (SGSS), and national death registry records from the Office for National Statistics (ONS). Vaccination status is available in the GP record directly via the National Immunisation Management System (NIMS). Healthcare worker status is recorded for all vaccine recipients at the time of vaccination, and this information is sent to OpenSAFELY-TPP from NHS Digital's COVID-19 data store.

## Eligiblity criteria

We used observational data to emulate a sequence of target trials @dagan2021 @barda2021, starting on consecutive days between `r format(as.Date(study_dates$studystart_date), "%d %B %Y")` and `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")` inclusive.

Eligibility on each day was as follows: adults (18 years and over) who were registered at a GP practice using TPP's SystmOne clinical information system; received a primary COVID-19 vaccination course of either two BNT162b2 doses or two ChAdOx1-S doses (mRNA-1273 was not considered due to small numbers); not a vaccinated health care worker, not a resident in a care or nursing home, and not medically housebound or receiving end-of-life care (due to distinct clustering of vaccine uptake and infection risk in these groups that could not be reasonably measured and controlled for); no evidence of SARS-CoV-2 infection or COVID-19 disease within 90 days prior to recruitment; not undergoing an unplanned hospital admission; information on sex, ethnicity, deprivation, and NHS region was complete.

Additional eligibility criteria, based on sufficiently many people being vaccinated on each day by region, priority group, and week of second dose. If the rolling weekly average count of BNT162b2 booster doses within these strata was below 50 then recruitment did not occur on that day. This helped to ensure that unboosted people were both eligible and able to receive a booster dose at the time of potential recruitment.

We excluded matched pairs from days in which fewer than 10% of eligible treated people were successfully matched. See supplementary materials for further details.

## Treatment groups

For each trial, the treatment group comprised people who received a BNT162b2 booster dose on that day, and the control group comprised the same number of randomly selected unboosted people, matched on key characteristics as follows: primary course vaccine type (BNT162b2 or ChAdOx1-S); calendar week of the second vaccine dose; NHS region; prior infection; immunosuppression; vaccine priority group (as determined by the Joint Committee for Vaccine and Immunisation expert working group). People selected as controls were not eligible to be included again as a control in a subsequent trial, but were eligible for selection into the treatment group of a subsequent trial.

## Outcomes

The following three outcomes were considered: documented SARS-CoV-2 infection; COVID-19 hospital admission; and COVID-19 death. SARS-CoV-2 infections were identified using SGSS testing records and based on swab date. Both polymerase chain reaction (PCR) and lateral flow tests are included, without differentiation between symptomatic and asymptomatic infection. COVID-19 hospital admissions were identified using HES in-patient hospital records with U07.1 or U07.2 reason for admission ICD-10 codes. COVID-19 deaths with the same COVID-19 ICD-10 codes mentioned anywhere on the death certificate (i.e., as an underlying or contributing cause of death) were included.

<!--# COVID-19 related A&E attendance and critical/intensive care admission we also considered but there were doubts over data the reliability of event ascertainment in these cases. -->
<!--# COVID-19 A&E attendances were identified using HES emergency care records with U07.1 ("COVID-19, virus identified") or U07.2 ("COVID-19, virus not identified") ICD-10 diagnosis codes @emergenc. -->

## Follow-up

Each matched pair was followed from the day of recruitment (time zero or baseline) until the earliest of death, GP practice de-registration, booster receipt by the control, 10 weeks, or `r format(as.Date(study_dates$studyend_date), "%d %B %Y")`.

@@

For each outcome, we estimated cumulative incidence curves using the Kaplan-Meier estimator separately in the treated and control groups.

We estimated period-specific hazard ratios for treatment versus control in consecutive two-week periods using a Cox model. Baseline hazards were allowed to differ by recruitment day, region, and week and type of second vaccine dose. The following variables were included as covariates: age, sex (male or female), English Index of Multiple Deprivation (IMD, grouped by quintile), ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census), NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West), number of conditions in the clinically "at risk" classification, as per national prioritisation guidelines, the number of SARS-CoV-2 tests in the 6 months prior to the study start date (via SGSS), evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation), learning disabilities, severe mental illness, undergoing an unplanned hospital admission, undergoing a planned hospital admission. A robust variance estimator was used to account for participant clustering.

<!-- [The primary estimand is the average treatment effect in the recruited population. To estimate this, expected cumulative incidence rates are estimated for each participant assuming treatment was received by everyone in both treatment groups at time zero. These curves are then averaged over all participants to give the marginal cumulative incidence. This is repeated assuming nobody is treated in either group, and their difference taken. Calculating standard errors for these quantities is challenging, whether approached analytically (due to complex variance components) or computationally (via bootstrapping, due to the large sample sizes involved) and are therefore not reported. ] -->

We additionally estimated booster effectiveness in subgroups defined by primary vaccine course (both BNT162b2 or both ChAdOx1-S) by repeating the analysis separately in each subgroup.

Booster effectiveness is defined as one minus the hazard ratio, and expressed as a percentage.

## Missing data

After exclusions for missing values on demographic variables as described earlier, there were no missing values in remaining variables as they were each defined by the presence or absence of clinical codes or events in the patient record. This was therefore a complete-case analysis.

@@

Data management and analyses were conducted in Python version 3.8.10 and R version 4.0.2. All code is available at <https://github.com/opensafely/booster-effectiveness>, and is shared openly for review and re-use under MIT open license. Codelists are available at <https://www.opencodelists.org/>. No person-level data is shared. Any reported figures based on counts below 6 are redacted or rounded for disclosure control.

@@

A total of `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_treated))` BNT162b2 booster doses were administered between `r format(as.Date(matchsummary_treated$firstrecruitdate), "%d %B %Y")` and `r format(as.Date(matchsummary_treated$lastrecruitdate), "%d %B %Y")` in adults registered at a TPP practice, with `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_eligible))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_eligible_treated))`) eligible for inclusion into the treatment group. Of these, `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` (`r label_percent(0.1)(filter(matchsummary, treatment=="pfizer") %>% pull(prop_matched_eligible))`) were successfully matched with unboosted controls (Figure 1).

@@

Participant characteristics at trial recruitment between treatment groups (Table 1) were typically well-balanced (necessarily so on matched characteristics), though with some exceptions, for instance unboosted controls had higher levels of deprivation, higher rates of learning disabilities and severe mental illness, and lower SARS-CoV-2 testing frequency. These unbalanced characteristics were adjusted for in the Cox regression models.

There were `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(fup_years))` person-years of follow-up across both treatment groups (Table 2).

@@

#### Figure 1: Booster vaccination uptake, treatment group eligibility and matching success

@@

xmin <- min(matchcoverage$vax3_date )
xmax <- max(matchcoverage$vax3_date )+1
matchcoverage  %>%
  filter(treatment=="pfizer") %>%
  mutate(
    # vax12_type = fct_case_when(
    #   vax12_type == "pfizer-pfizer" ~ "BNT162b2",
    #   vax12_type == "az-az" ~ "ChAdOx1-S",
    #   vax12_type == "moderna-moderna" ~ "mRNA-1273",
    #   TRUE ~ NA_character_
    # ),
    course = fct_case_when(
      vax12_type=="pfizer-pfizer" & treatment=="pfizer" ~ "2× BNT162b2  / 1× BNT162b2",
      vax12_type=="az-az" & treatment=="pfizer" ~ "2× ChAdOx1-S / 1× BNT162b2",
      vax12_type=="pfizer-pfizer" & treatment=="moderna" ~ "2× BNT162b2 / 1× mRNA-1273",
      vax12_type=="az-az" & treatment=="moderna" ~ "2× ChAdOx1-S / 1× mRNA-1273",
      TRUE ~ NA_character_
    )
  ) %>%
  ggplot() +
  geom_col(
    aes(
      x=vax3_date+0.5,
      y=cumuln,
      group=fct_rev(status),
      fill=fct_rev(status),
      colour=NULL
    ),
    position=position_stack(),
    alpha=0.8,
    width=1
  )+
  geom_rect(xmin=xmin, xmax= xmax+1, ymin=0, ymax=6, fill="grey", colour="transparent")+
  facet_wrap(vars(course))+
  scale_x_date(
    breaks = unique(lubridate::ceiling_date(matchcoverage$vax3_date, "1 month")),
    limits = c(lubridate::floor_date((xmin), "1 month"), NA),
    labels = scales::label_date("%d/%m"),
    expand = expansion(add=1),
  )+
  scale_y_continuous(
    labels = scales::label_number(accuracy = 1, big.mark=","),
    expand = expansion(c(0, NA))
  )+
  scale_fill_brewer(type="qual", palette="Set2", direction = -1)+
  scale_colour_brewer(type="qual", palette="Set2", direction = -1)+
  labs(
    x="Date",
    y="Cumulative booster vaccines",
    colour=NULL,
    fill=NULL,
    alpha=NULL
  ) +
  theme_minimal()+
  theme(
    axis.line.x.bottom = element_line(),
    axis.text.x.top=element_text(hjust=0),
    strip.text.y.right = element_text(angle = 0),
    axis.ticks.x=element_line(),
    legend.position = "bottom"
  )+
  NULL
cat("  \n\n")
```

@@

Participant characteristics as on the day of recruitment into the treatment or control group.

@@

#### Table 2: Follow-up and outcomes

Total follow-up and incidence rates for each outcome, by treatment group.

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath"),
    treatment == "pfizer"
  ) %>%
  left_join(
    matchsummary %>% select(treatment, n_eligible, prop_matched_eligible),
    by=c("treatment")
  ) %>%
  mutate(
    treatment_decr = paste0(
      treatment_descr, " - ",
      label_number(1, big.mark=",")(n_eligible), "eligible, ",
      label_number(1, big.mark=",")(n_1), "matched (",
      label_percent(0.1)(prop_matched_eligible), ")."
    )
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rrECI
  ) %>%
  mutate(
    rate_1=if_else(rate_1<0.001, "<0.001", label_number(0.001)(rate_1)),
    rate_0=if_else(rate_0<0.001, "<0.001", label_number(0.001)(rate_0)),
  )
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

    rate_0 = "Incidence rate",
    rate_1 = "Incidence rate",
    rrECI = "Incidence rate ratio (95% CI)"
  ) %>%
  tab_spanner(
    label = "Boosted",
    columns = ends_with("1")
  ) %>%
  tab_spanner(
    label = "Unboosted",
    columns = ends_with("0")
  ) %>%
  fmt_missing(
    everything(),
    missing_text="--"
  ) %>%
  cols_align(
    align = "right",
    columns = everything()
  )

@@

Participants in the emulated trial contributed `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="postest", fup_period=="All") %>% pull(events))` documented infection, `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="covidadmitted", fup_period=="All") %>% pull(events))` COVID-19 hospital admissions, and `r label_number(1, big.mark=",")(filter(incidences, treatment=="pfizer", outcome=="coviddeath", fup_period=="All") %>% pull(events))` COVID-19 deaths (Table 2).
